A Practical Optional Type System for Clojure

### 2012-10-19 dissertation/implementation.tex
Intro to variable-arity

A substitution maps type variables to an upper and lower bound. 
It is valid to replace all occurrences of the type variables named in the substitution 
with a type between their associated bounds.
For example, usually the substitution is applied to \lstinline|R| by the caller of \lstinline|infer|.
Whether to substitute the upper or lower bound is determined by the variance
of the type variable in relation to the type being substituted. 
The specific algorithm is described in Local Type Inference.

@@

F-bounded polymorphism allows type variables to refer to themselves in their bounds.

@@

It also required changes to the polymorphic type inference algorithm.
The constraint-generation portion

@@

@@

\section{Filtering on Negative Propositions}
\label{sec:filterneg}

>>>>

A substitution maps type variables to a type.
It is valid to replace all occurrences of the type variables named in the substitution 
with their associated type.
For example, substitution are often applied to \lstinline|R| by the caller of \lstinline|infer|
to eliminate the type variables in \lstinline|X| and \lstinline|Y|.

@@

F-bounded polymorphism allows type variables to refer to themselves in their type bounds.
Bounds are checked after a substitution is generated, guaranteeing no substitution
can violate type variable bounds. To support F-bounds, the substitution being checked is applied to
the lower and upper bounds for each type variable, and the type associated with the type variable 
in the substitution is checked to be between these bounds.

@@

It also required changes to the polymorphic type inference, where each
reference to a type variable required a special case for a dotted type variable.
For example, the constraint-generation algorithm for Local Type Inference
features extra kinds of constraints for dotted variables.
Strickland, Tobin-Hochstadt, and Felleisen elaborate on the particular changes
required for the Typed Racket implementation\cite{STF09}.

@@

\section{Proposition Inference for Functions}
\label{sec:filterneg}

A feature not yet implemented in Typed Racket is the ability
to infer new propositions based on existing propositions of a function.
This feature was added to Typed Clojure to support filtering
a sequence based on negative information, such as filtering values
that are \emph{not} \lstinline|nil|.

\begin{lstlisting}[caption=Type annotation for filter, label=lst:filterann]
(ann clojure.core/filter 
  (All [x y]
    [[x -> Any :filters {:then (is y 0)}] (U nil (Seqable x)) -> (Seqable y)]))
\end{lstlisting}

To better understand the problem, listing \ref{lst:filterann} presents the type of \lstinline|filter|,
which takes a function \lstinline|f| and a sequence \lstinline|s| as arguments, and returns a sequence that 
contains each element in \lstinline|s| such that applying \lstinline|f| to the element returns a true value.
The \lstinline|:filters| syntax requires some explanation. 
Function types support an optional \emph{filter set} attached to its return type, written as a map
with \lstinline|:then| and/or \lstinline|:else| keys (if omitted, they default to the trivially true proposition which has no effect). 
The ``then-proposition'' and ``else-proposition'' are added
to the type environment when the return value is a true and false value respectively.
For example, the filter set \lstinline|{:then (is y 0)}| is read ``if the return value is a true value, then the first argument
must be of type y, otherwise if it is a false value, nothing interesting is enforced''
The type given for \lstinline|filter| works because the type variable
\lstinline|y| occurs in both the ``then-proposition'' of the first argument and the return type 
\lstinline|(Seqable y)|.

\begin{lstlisting}[caption=Troublesome filter, label=lst:filtertrouble]
(filter (fn [a] (not (nil? a))) coll)|
\end{lstlisting}

The difficulty starts with something like listing \ref{lst:filtertrouble},
where the inferred filter set for the first argument to \lstinline|filter|
is \lstinline|{:then (! nil 0) :else (is nil 0)}|\footnote{(! nil 0) is the proposition that the first argument is \emph{not}
of type \lstinline|nil|.}. The ``then-proposition'' \lstinline|(! nil 0)| does not fit with
\lstinline|(is y 0)| that \lstinline|filter| expects.

We can sometimes get around this if we already have a predicate with a positive
``then-proposition''. For example, if we are filtering out \lstinline|nil| values
from a sequence of type \lstinline|(Seqable (U Number nil))|, we can replace
listing \ref{lst:filtertrouble} with \lstinline|(filter number? coll)|,
where \lstinline|number?| has the filter set \lstinline|{:then (is Number 0) :else (! Number 0)}|.
This does not work, however, when filtering a sequence of type like \lstinline|(Seqable (U x nil))|
for some unspecified \lstinline|x| because there is no built-in predicate with ``then-proposition''
\lstinline|(is x 0)|.

\begin{lstlisting}[caption=Filtering with negative propositions, label=lst:filtergood]
  (filter (ann-form 
            #(not (nil? %))
            [(U nil x) -> boolean :filters {:then (is x 0)}])
          mvs)
\end{lstlisting}

Instead, we generate new propositions using a technique 
suggested by Tobin-Hochstadt\cite{Tob12},
that follows from the occurrence typing calculus defined 
by Tobin-Hochstadt and Felleisen\cite{TF10}.
First the filter set for functions are inferred as usual.
Then the ``then-proposition'' is applied to the environment and new propositions
are added to the filter set for each variable in the type environment that changes.
The same procedure is followed for the ``else-proposition''.

Using this technique, the anonymous function in listing 
\ref{lst:filtergood}\footnote{The Clojure syntax \lstinline|#(not (nil? \%))| 
is equivalent to \lstinline|(fn [a] (not (nil? a)))|}
has the filter set \lstinline|{:then (& (! nil 0) (is x 0)) :else (is nil 0)}|
which is good enough to infer the \lstinline|filter|ed result as \lstinline|(Seqable x)|.

@@

### 2012-10-19 dissertation/intro.tex
Intro to variable-arity

% Type Systems for lisp?
% Why type systems?
% Why Clojure?

@@

Recently languages have been created to or been modified to support aspects
of both static and dynamic typing.
Dart\footnote{http://www.dartlang.org/} is dynamically typed but offers optional type annotations
that do not affect runtime semantics. 
Typescript\footnote{http://www.typescriptlang.org/} adds an optional type system to Javascript,
a well-known dynamic language.

@@

Typed Clojure supports F-bounded polymorphism. All type variables may have
upper and lower bounds.

@@

@@

Where particularly object-oriented languages reach for objects, Clojure
utilises maps. Clojure provides a map literal using curly braces. For example,
\lstinline|{:a 1, :b 2}| is a map with two key-value entries: from Keyword key \lstinline|:a|
to value \lstinline|1|, and Keyword key \lstinline|:b| to value \lstinline|2|. Note that commas are always
whitespace in Clojure and are included for readability.

Typed Clojure provides a heterogeneous map type, restricted to 
maps with singleton Keyword keys. This restriction is reflected
in the syntax for defining heterogeneous map types.

\begin{lstlisting}[caption=Heterogeneous map types in Typed Clojure]
...
(ann config
     (HMap {:file String,
            :ns Symbol}))
(def config
  {:file "clojure/core.clj",
   :ns 'clojure.core})
...
\end{lstlisting}

@@

\subsection{Multi-Arity Functions}

Clojure function objects support multiple arities, allowing dispatch
on the number of supplied arguments.

\begin{lstlisting}[caption=Annotating multi-arity functions in Typed Clojure]
(ns typed.test.poly
  (:require [typed.core :refer [ann AnyInteger check-ns cf]])
  (:import [clojure.lang Seqable]))

(ann repeatedly'
     (All [x]
       (Fn [[-> x] -> (Seqable x)]
           [AnyInteger [-> x] -> (Seqable x)])))
(defn repeatedly'
  "Takes a function of no args, presumably with side effects, and
  returns an infinite (or length n if supplied) lazy sequence of calls
  to it"
  ([f] (lazy-seq (cons (f) (repeatedly' f))))
  ([n f] (take n (repeatedly' f))))
\end{lstlisting}

This example defines \lstinline|repeatedly'|\footnote{Identical to \lstinline|clojure.core/repeatedly|},
a multi-arity function taking either a zero-argument function, or an integer and a zero-argument function as arguments.
The type variable \lstinline|x| links the return type of the generator function to
the resulting \lstinline|Seqable| instance.

When checking the definition of multi-arity functions, Typed Clojure
finds the matching function types for a particular arity by number of arguments
and requires each relevant type to check successfully.

@@

Clojure functions can take an optional \emph{rest} argument.

@@

\subsubsection{Non-uniform Variable-Arity}

@@

Typed Clojure uses occurrence typing to improve inference\footnote{TODO occurrence typing reference}.
Occurrence typing is used to refine types for immutable local bindings 
as programs follow conditional branches.

\begin{lstlisting}[caption=Example of occurrence typing in Typed Clojure]
(ann num-vec2 
     [(U nil Number) (U nil Number) -> (Vector* Number Number)])
(defn num-vec2 [a b]
  [(if a a 0) 
   (if b b 0)])
\end{lstlisting}

@@

Typed Clojure currently provides safe interop\footnote{Interop is short for interoperability} with (non-Generic) Java
\footnote{Compatibility with Generic Java is planned future work}.

The key to safe Java interop is the treatment of Java's \emph{null}.
\emph{null} is a subtype to all reference types
represented in Clojure by \lstinline|nil|. Unlike Java's type system
Clojure explicitly separates \emph{null} and reference types, allowing
Typed Clojure to express
\emph{nullable}
\footnote{A type is \emph{nullable} if it may also be an instance of \emph{null},
which is expressed in Typed Clojure by creating a union of the reference type and \lstinline|nil|.}
positions.

>>>>

@@

Recently languages have been created to or been modified to support aspects
of both static and dynamic typing.
Dart\footnote{http://www.dartlang.org/} is dynamically typed but offers a simple form of optional
static typing that specifically do not affect runtime semantics.
Typescript\footnote{http://www.typescriptlang.org/} adds an optional type system to Javascript,
a well-known dynamic language.

@@

Typed Clojure supports F-bounded polymorphism, first introduced by Canning, Cook, Hill and Olthoff\cite{CCHOM89}. 
F-bounded polymorphism is an extension of bounded polymorphism, where polymorphic type variables
can be restricted by \emph{bounds}.
In particular, F-bounded polymorphism allows type variable bounds to recursively refer to the
variable being bounded. Typed Clojure supports upper and lower type variable bounds.

@@

\subsection{Singleton Types}

Following Typed Racket, singleton types for certain values are provided
in Typed Clojure.
A singleton type is a type with a single member, like \lstinline|1|,
\lstinline|:a|, or \lstinline|"a"|\footnote{Strings are delimited by " in Clojure}.
Typed Clojure provides syntax for singleton types, either by passing
the value literal to the \lstinline|Value| primitive, or by prefixing
a quote (').

\begin{lstlisting}[caption=Singleton Types, label=lst:singletoneeg]
(ann k ':my-keyword)
(def k :my-keyword)
\end{lstlisting}

Listing \ref{lst:singletoneeg} shows a simple example of using
singleton types in Typed Clojure.
Singleton types are are discussed further in section \ref{sec:intersectionunion}.

@@

A novel feature of Clojure compared to other dialects of Lisp (including Racket)
is an emphasis on key-value paired, unordered \emph{maps}.
Where particularly object-oriented languages reach for objects, Clojure
utilises maps. Clojure provides a map literal using curly braces. For example,
\lstinline|{:a 1, :b 2}| is a map value with two key-value entries: from keyword key \lstinline|:a|
to value \lstinline|1|, and keyword key \lstinline|:b| to value \lstinline|2|. Note that commas are always
whitespace in Clojure and are included occasionally for readability.

Typed Clojure provides a heterogeneous map type which captures this common
``maps as objects'' pattern. A heterogeneous map type has only \emph{positive}
information on the types of key-value entries. In other words, it conveys
whether a particular key is present, but not whether it is absent.
The implications of this is discussed in section \ref{ref:designhmap}.
Heterogeneous maps only support keys that are singleton Keyword types. This restriction is reflected
in the syntax for defining heterogeneous map types.

\begin{lstlisting}[caption=Heterogeneous map types in Typed Clojure]
...
(ann config '{:file String, :ns Symbol}))
(def config
  {:file "clojure/core.clj",
   :ns 'clojure.core})
...
\end{lstlisting}

@@

@@

Functions in Clojure are multi-arity, which means a function
can be defined with several function bodies (or \emph{arities})
and which arity is executed depends on the number of arguments passed
to the function. A function can have any number of arities with fixed parameters, and at most one arity with
variable-parameters. Each arity must have a unique number of
parameters and have a lower number of parameters than the ``variable arity'', if present.

Strickland, Tobin-Hochstadt, and Felleisen invented a calculus and corresponding
implementation in Typed Racket for variable-arity polymorphism \cite{STF09}
that is sufficient to handle uniform and non-uniform variable-arity functions.
Typed Clojure includes an implementation of the most immediately useful parts of variable-arity
polymorphism using algorithms, nomenclature, and implementation based on this work.

@@

A function with uniform variable parameters can treat its variable parameter
as a homogeneous list. 
Strickland et al. attaches a \emph{starred pre-type} \lstinline|T *| to the right of the fixed arguments
in a function type, where \emph{T} is some type, and we take an identical approach.
For example, \lstinline|+| in Clojure accepts any number of arguments
of type \lstinline|Number|, represented by the type \lstinline|[Number * -> Number]|.

\begin{lstlisting}[caption=Typing multi-arity functions, label=lst:noteq]
(ann not= (Fn [Any -> boolean]
              [Any Any -> boolean]
              [Any Any Any * -> boolean]))
(defn not=
  "Same as (not (= obj1 obj2))"
  ([x] false)
  ([x y] (not (= x y)))
  ([x y & more]
   (not (apply = x y more))))
\end{lstlisting}

It is common to find Clojure library functions that define seemingly redundant
function arities for performance reasons.
Listing \ref{lst:noteq} defines the multi-arity function \lstinline|not=|,
taken from the Clojure standard library \lstinline|clojure.core| that uses this pattern.
\lstinline|not=| has three arities in its definition, including
one that takes a variable number of arguments.
The \lstinline|Fn| type constructor builds an \emph{ordered function intersection type} from
function types. Each arity must have at least one matching function type associated with it.
The two ``fixed arities'' are given familiar function types, with one and two fixed parameters respectively.
The type given for the arity with variable arguments \lstinline|[Any Any Any * -> boolean]|
uses a starred pre-type to signify any number of arguments of type \lstinline|Any|
can be provided to the right of its fixed arguments.

\subsubsection{Non-uniform Variable-Arity Functions}

Where \emph{uniform} variable-arity function types use \emph{starred pre-types}, \emph{non-uniform}
variable-arity function types use \emph{dotted pre-types}.
Typed Clojure supports usages of \emph{non-uniform} variable-arity functions,
where the variable parameter is a heterogeneous list, represented by a \emph{dotted type variable}.

For example, the variable argument function \lstinline|clojure.core/map| takes a function and one or more sequences,
and returns the result of applying the function argument to each element of the sequences in a pair-wise fashion.
Its type is given in listing \ref{lst:maptype}.

\begin{lstlisting}[caption=Type signature for \lstinline|clojure.core/map|, label=lst:maptype]
(ann clojure.core/map
     (All [c a b ...]
       [[a b ... b -> c] (U nil (Seqable a)) (U nil (Seqable b)) ... b -> (LazySeq c)]))
\end{lstlisting}

By adding \lstinline|...| after the last type variable in an \lstinline|All| binder
we can introduce a \emph{dotted type variable} into scope, which is a placeholder for a sequence of types.
A \emph{dotted pre-type} \lstinline|T ... b| over \emph{base} \lstinline|T| (a type) and \emph{bound}
\lstinline|b| (a dotted type variable)
serves as a placeholder for this sequence of types.
Dotted pre-types must appear to the right of all fixed parameters in a function type,
and cannot be mixed with other kinds of variable parameters like starred pre-types.
When a sequence of types of length \emph{n} is associated with a dotted pre-type, 
the dotted pre-type is expanded to \emph{n} copies of \lstinline|T|.
One other special property of a dotted pre-type is that the bound \lstinline|b|
is \emph{in scope} as a \emph{normal} type variable in its base \lstinline|T|.

To demonstrate dotted pre-types, we use \lstinline|typed.core/inst| to instantiate
\lstinline|map|. \lstinline|inst| takes a polymorphic expression and a number of types
that are satisfactory for instantiating the polymorphic type and returns an expression
of the instantiated polymorphic type.

The instantiation

\begin{lstlisting}
(inst map Number boolean String)
\end{lstlisting}

returns an expression of type

\begin{lstlisting}
[[boolean String Integer -> Number] (U (Seqable boolean) nil) (U (Seqable String) nil) (U (Seqable Integer) nil) -> (LazySeq Number)]
\end{lstlisting}

However, if sufficient types are given, the instantiation for \lstinline|map| can be inferred.
The invocation

\begin{lstlisting}
(map (ann-form (fn [a b c] c) 
               [boolean String Number -> Number]) 
     [true false] ["mystr" "astr"] [1 4])
\end{lstlisting}

uses \lstinline|typed.core/ann-form| to assign an expected to
the first argument,
which is sufficient to infer the result type

\begin{lstlisting}
(LazySeq Number)
\end{lstlisting}

@@

It is common in Clojure, like other dynamically-typed languages, to
encode implicit type invariants in conditional tests.
In listing \ref{lst:numvec2} conditional tests are used
to refine the type of the bindings \lstinline|a| and \lstinline|b|.
Occurrence typing is a technique useful for capturing these kinds
of type invariants (occurrence typing is discussed in further detail in section \ref{sec:OccurrenceTyping}).

\begin{lstlisting}[caption=Example of occurrence typing in Typed Clojure, label=lst:numvec2]
(ann num-vec2 
     [(U nil Number) (U nil Number) -> (Vector* Number Number)])
(defn num-vec2 [a b]
  [(if a a 0) 
   (if b b 0)])
\end{lstlisting}

@@

Typed Clojure currently provides safe interoperability
\footnote{Compatibility with Generic Java is planned future work}.

The key to safe Java interoperability is the treatment of Java's \emph{null}.
\emph{null} is a subtype to all reference types
represented in Clojure by \lstinline|nil|. Unlike Java's type system
Clojure explicitly separates \emph{null} and reference types, allowing
Typed Clojure to express
\emph{nullable}
\footnote{A type is \emph{nullable} if it may also be an instance of \emph{null},
which is expressed in Typed Clojure by creating a union of the reference type and \lstinline|nil|.}
positions.

### 2012-10-19 dissertation/lit-review.tex
Intro to variable-arity

\section{Intersection and Union Types}

Intersection and union types are interesting type constructs relevant to capturing the complicated
types common in dynamic languages.
An intersection of types \lstinline|a| and \lstinline|b|, written \lstinline|(I a b)| in Typed Clojure,
can be used safely in both positions expecting type \lstinline|a|,
and positions expecting type \lstinline|b|.
Intuitively, \lstinline|(I a b)| is both of type \lstinline|a| \emph{and} \lstinline|b|.
A union of types \lstinline|a| and \lstinline|b| is \emph{either} of type \lstinline|a| or
\lstinline|b|, written \lstinline|(U a b)| in Typed Clojure,
and can be used safely in positions expecting a type that is \emph{both} type \lstinline|a|
and \lstinline|b|.
For example, \lstinline|(U Number Boolean)| cannot be used in positions expecting \lstinline|Number|.

@@

SML CIDRE is a refinement type checker for Standard ML by Davies \cite{Dav05}
that features a form of intersection types. It is used for function overloading
and checking invariants involving refinement types.
\begin{verbatim}
Freeman & Pfenning, Refinement types
\end{verbatim}

St-Amour, Tobin-Hochstadt, Flatt, and Felleisen
describe the \emph{ordered intersection types} used in Typed Racket\cite{St12}
that provide a kind of function overloading.
Typed Clojure takes a similar approach
\begin{verbatim}
described in chapter
\end{verbatim}
Typed Racket also uses union types

>>>>

\section{Intersection, Union, and Singleton Types}
\label{sec:intersection_types}

% Hiyashi Singleton types
% Intersection types - CDV81
% Unions in soft typing CF91
% - Static Type Inference in a Dynamically Typed Lanugage
%  - cited by CF91, which has unions

Intersection and union types are interesting type constructs relevant to capturing the complicated
types common in dynamic languages.
An expression of type \lstinline|(I a b)|, an intersection type including types \lstinline|a|
and \lstinline|b| in Typed Clojure, can be used safely in both positions expecting type \lstinline|a|,
and positions expecting type \lstinline|b|.
An expression of type \lstinline|(U a b)|, a union type including types \lstinline|a| in Typed Clojure,
and expressions of this type can be used safely in positions 
that expect a type that is \emph{either} type \lstinline|a|
or \lstinline|b|.
For example, \lstinline|(U Number Boolean)| cannot be used in positions expecting \lstinline|Number|.
% TODO Intersection example

@@

Refinement types add a level of types refining an existing type system
in order to type check more detailed properties and invariants than standard static type systems
(described by Freeman and Pfenning\cite{FP91}).
Intersection types are critical here to allow more than one property or invariant
to be expressed for a function.
SML CIDRE is a refinement type checker for Standard ML by Davies \cite{Dav05}.
\begin{verbatim}
Refinement types are similar to this project ... how?
\end{verbatim}

St-Amour, Tobin-Hochstadt, Flatt, and Felleisen
describe the \emph{ordered intersection types} used in Typed Racket\cite{St12}
that provide a kind of function overloading.
Typed Clojure takes a similar approach
\begin{verbatim}
described in chapter
\end{verbatim}
Typed Racket also uses disjoint union types,
